---
title: "Straight lines"
output:
  pdf_document: default
  html_notebook: default
---

# Looking for straight line relationships

We go back to the AirBnB data:

## Loading the data

```{r}
airbnb <- read.csv('amsterdam_airbnb_2_2021.csv')
head(airbnb)
```

There are some very high values in the `price` column.

```{r}
hist(airbnb$price)
```

We subset the data frame to give us just the more realistic prices.

```{r}
ok_airbnb <- subset(airbnb, airbnb$price < 1000)
```

Now let's try again:

```{r}
hist(ok_airbnb$price)
```

Now we want to look at relationship between the number of bedrooms and the price.

First, to be nice and clean, I select the columns I want:

```{r}
price <- ok_airbnb$price
beds <- ok_airbnb$bedrooms
```

I am interested in the relationship between number of bedrooms and price:

```{r}
plot(beds, price)
```

It looks like there might be some kind of straight-line relationship between the number of bedrooms and the price.  Here's what R thinks that is.  Remember the tilde `~` means "as a function of"

```{r}
fit <- lm(price ~ beds)
summary(fit)
```

The cooefficients of the fit give the intercept and slope:

```{r}
fit$coefficients
```

```{r}
intercept <- fit$coefficients[1]
slope <- fit$coefficients[2]
intercept
slope
```

Put that straight line on the plot:

```{r}
plot(beds, price)
abline(intercept, slope)
```

Actually, `abline` is clever enough to work out that is what we mean, when we pass the `fit` itself:

```{r}
plot(beds, price)
abline(fit)
```

The slope is the increase in price for an increase by one bedroom.

*Correlation* is a measure of how well this line explains the data.  It varies continuously between -1 and 1, where 1 and -1 mean the line fully explains the data, and zero means the line explains nothing in the data. Positive correlations mean the slope is positive, negative mean the slope is negative.

```{r}
cor(beds, price)
```

Note that `price` and `beds` have very different spread and center.

```{r}
hist(price)
```

```{r}
hist(beds)
```

We will use the mean and standard deviation as measures of center and spread:

```{r}
mean(price)
sd(price)
```

```{r}
mean(beds)
sd(beds)
```

The distributions have very different centers (means) and spreads (standard deviations).

We can force them to have the same mean and standard deviation by applying the z-score transformation.

## Z score transformation

We first subtract the mean, to give the deviations.  The sum and therefore the mean of the deviations, is (almost exactly) zero:

```{r}
dev_price <- price - mean(price)
mean(dev_price)
```

The standard deviation hasn't changed though:

```{r}
sd(dev_price)
```

We set the values to have a standard deviation of 1, by dividing by the standard deviation.  This gives the *z score*:

```{r}
z_price <- dev_price / sd(price)
head(z_price)
```

Now the values have a standard deviation of 1:

```{r}
sd(z_price)
```

We can see the center and spread on the histogram:

```{r}
hist(z_price)
```

We do the same for `beds`:

```{r}
dev_beds <- beds - mean(beds)
z_beds <- dev_beds / sd(beds)
head(z_beds)
```

```{r}
hist(z_beds)
```

The plot of `z_beds` and `z_price` looks similar to the original plot, but now the ranges on the axes are comparable:

```{r}
plot(z_beds, z_price)
```

We find the "best-fitting" line (we haven't defined what we mean by "best" yet).

```{r}
z_fit <- lm(z_price ~ z_beds)
summary(z_fit)
```

Notice the Intercept is (almost exactly) zero.

Now look at the slope:

```{r}
z_slope <- z_fit$coefficients[2]
z_slope
```

Notice this is exactly the correlation of the original variables:

```{r}
cor(beds, price)
```

Another way of explaining the correlation is to say it is the slope of the line relating two variables, once they have been converted to z-scores.

```{r}
plot(z_beds, z_price)
abline(z_fit$coefficients[1], z_fit$coefficients[2])
```

Because it is the slope of the line, it means that for every 1 unit increase in `z_beds`, `z_price` goes up by 0.55 units.

Put another way, for every increase of 1 standard deviation in `beds`, `price` increases by 0.55 standard deviations.

How do we find this slope?

We can ask R, of course, as we have done.  We can also calculate it simply ourselves.

First we multiply the z scores together, z score for position 1 times z score for position 1, z score for position 2 times z score for position 2, and so on:

```{r}
z_beds[1] * z_price[1]
z_beds[2] * z_price[2]
z_beds[3] * z_price[3]
```

We can do this all in one go with:

```{r}
z_by_z <- z_beds * z_price
head(z_by_z)
```

We take the sum of these values and divide by (n - 1).  It's the correlation, and the slope!

```{r}
n <- length(z_by_z)
adj_mean_zz <- sum(z_by_z) / (n - 1)
adj_mean_zz
```

We said before that this means there is a 0.55 standard deviation increase in `price` for every 1 standard deviation increase in `beds`.

Put another way, this means there is an increase of `0.55 * sd(price)` on the y axis for every increase of `sd(beds)` on the x axis. Here's the increase on the y axis for a one standard deviation increase in x:

```{r}
for_1_sd_beds <- z_slope * sd(price)
for_1_sd_beds
```

For an increase of 1 on the original x axis (`beds`), we divide by `sd(beds)`:

```{r}
for_1_bed <- for_1_sd_beds / sd(beds)
for_1_bed
```

This is the slope of the original line, before we did the z transformation.  We stored that before as `slope`:

```{r}
slope
```

## Recipe for the correlation coefficient

* Convert both sets of values to z scores
* Multiply the z score values together
* take the sum, divide by (n - 1)
* call this `r`.

## Getting the "best-fit" slope for the original data

* Multiply `r` by the standard deviation of the the values on the y axis and divide by the standard deviation of the values on the x axis.

## Getting the intercept for the original data

We know the slope for the original data, after we have undone the division by the standard deviation that we used in the z-score procedure. Now we need to undo the subtraction of the means.  Doing this moves the coordinate from the z-score graph to (0, 0) to position (mean of x, mean of y) on the original axes.   Now we have a point on the original line (mean of x, mean of y).  Given this point, and the slope, we can calculate the intercept.  To see why, draw a picture of the point, and the slope, and see where the line crosses the y axis.

```{r}
mean(price) - slope * mean(beds)
```

Indeed, this is the same as:

```{r}
fit$coefficients[1]
```

## Another way

It may be useful to think of the steps above like this.

Call the original values `x` and `y`:

```{r}
x <- beds
y <- price
```

`n` is the number of values in each vector:

```{r}
n <- length(x)
```

Call the versions with their means subtracted `xm` and `ym`:

```{r}
xm <- x - mean(x)
ym <- y - mean(y)
```

Call the z score versions `xz` and `yz`:

```{r}
xz <- xm / sd(x)
yz <- ym / sd(y)
```

We found the "best" slope relating `xz` and `yz`:

```{r}
z_slope <- sum(xz * yz) / (n - 1)
z_slope
```

Remember that the intercept is 0.  This means that our line has the formula `yz = xz * z_slope`.

Because of our definition of `yz` and `xz` above, we can also write this as: `ym / sd(y) =  xm / sd(x) * z_slope`.

So `ym = xm * z_slope / sd(x) * sd(y)`.

For neatness define `s =  z_slope / sd(x) * sd(y)`.  In our case:

```{r}
s <- z_slope / sd(x) * sd(y)
s
```

Now we can write the above as ``ym = xm * s`.

Because of our definition of `xm` and `ym` above, we can also write:

`y - mean(y) = (x - mean(x)) * s`

So:

`y = x * s - mean(x) * s + mean(y)`

The slope is `s` and the intercept is `mean(y) - mean(x) * s`.   In our case:

```{r}
intercept <- mean(y) - mean(x) * s
intercept
```
